---
title: "CDS 492 Capstone Project"
author: "Manu Balasubramanian"
documentclass: article
geometry: margin=1in
fontsize: 11pt
output:
  pdf_document:
    toc: false
    df_print: kable
    fig_caption: false
    number_sections: false
    dev: pdf
    highlight: tango
  html_document:
    theme: default
    self_contained: true
    toc: false
    df_print: kable
    fig_caption: false
    number_sections: false
    smart: true
    dev: svg
---

```{r setup, include = FALSE}
# DO NOT ALTER THIS CHUNK
# Set knitr options
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  fig.width = 5,
  fig.asp = 0.618,
  out.width = "70%",
  dpi = 120,
  fig.align = "center",
  cache = FALSE
)

suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(infer))
suppressPackageStartupMessages(library(modelr))
suppressPackageStartupMessages(library(broom))

crime <- SPD_Officer_Involved_Shooting_OIS_Data 
```

```{r}
  crime
```
- This visualizes the crime data set in full


```{r}
service_summary <- crime %>%
  select(
    Years_of_SPD_Service,
    Officer_Gender,
    Rank,
    Officer_Race,
    Officer_Injured,
    Subject_Race,
    Subject_Gender,
  )
service_summary
```
- service_summary is a subset of the main dataset with only specific columns that will be modified and plotted


```{r}
service_summary %>%
  summarize(
    rows = n(),
    mean = mean(Years_of_SPD_Service, na.rm = TRUE),
    median = median(Years_of_SPD_Service, na.rm = TRUE),
    range = max(Years_of_SPD_Service, na.rm = TRUE),
    std = sd(Years_of_SPD_Service, na.rm = TRUE),
    iqr = IQR(Years_of_SPD_Service, na.rm = TRUE)
  )
```
-A summary of the years of service column is calculated using the summarize function

```{r}
table(crime$Subject_Gender)
table(crime$Officer_Gender)
table(crime$Officer_Injured)
table(crime$Officer_Race)
table(crime$Subject_Race)
```
- Summary statistics of the categorical variables are also calculated


```{r}
ggplot(service_summary, aes(x = Officer_Race)) + geom_bar() + theme(axis.text.x=element_text(size=5)) + labs(title = "Race of Officers", x = "Officer Race", y = "Count")

ggplot(service_summary, aes(x = Subject_Race)) + geom_bar() + theme(axis.text.x=element_text(size=5)) + labs(title = "Race of Subjects", x = "Subject Race", y = "Count")

ggplot(service_summary, aes(x = Officer_Gender)) + geom_bar() + theme(axis.text.x=element_text(size=5)) + labs(title = "Gender of Officers", x = "Officer Gender", y = "Count")

ggplot(service_summary, aes(x = Subject_Gender)) + geom_bar() + theme(axis.text.x=element_text(size=5)) + labs(title = "Gender of Subjects", x = "Subject Gender", y = "Count")
```
- This other visualization above now shows the spread of subject race. This gives us more information as there is a large increase in people of color compared to the race of the officers specifically those of African American descent. Like the previous visualizations, this is just a preliminary graphic that gives us the summary of race values for subjects in this dataset.

- This visualization above shows the spread of officer race values across this dataset. The majority of officers recorded in this dataset can be seen as white with very little officers of color. Even though there are mostly white officers, this does not show any form of racism as it does not show whether or not anyone was injured and it is just a general summary of all officers in this dataset.

- These visualizations above are the initial summary visualizations, visualizations below are apart of data analysis

```{r}
ggplot(service_summary, aes(x = Officer_Race)) + geom_bar() + facet_wrap(~ Officer_Gender) + theme(axis.text.x=element_text(size=3)) + labs(title = "Race and Gender of Officers")

ggplot(service_summary, aes(x = Subject_Race)) + geom_bar() + facet_wrap(~ Subject_Gender) + theme(axis.text.x=element_text(size=4)) + labs(title = "Race and Gender of Subjects")

```

- These plots are similar to those above where it plots the race of each officers and subjects but the gender column is faceted over the data to provide the race summary statistics along with gender. It can be seen that there were only reports of White, Black, and Asian females in this data set whereas there are instances of almost every race for males.


```{r}
fatal_summary <- crime %>%
  select(
    Fatal,
    Disposition,
    Officer_Race,
    Officer_Gender
  )
```

- fatal_summary takes specific columns to be used for calculations and modeling


```{r}
fatal_justified_table <- table(fatal_summary$Fatal, fatal_summary$Disposition)
barplot(fatal_justified_table, legend.text = TRUE, beside = TRUE, cex.names = 0.75, xlab = "Dispositions and Fatality", ylab = "Count", main = "Are Some Fatal Police Interactions Justified?")
```

- This plot is created using a table which compiles the total number of fatal incidents along with the officer race summary statistics. By analyzing the visualization above, most fatal incidents occur when the officer was within policy. This can be identified as an anomaly as typically, officers that are operating within policy would be assumed to be operating in a safe environment but according to the graphic, most fata interactions occur when officers are within policy. It can also be seen that when out of policy, there were no fatal incidents and many fatal incidents also happened when the subject was missing. When the stop was justified, there are more non-fatal encounters compared to fatal encounters. Although race and gender have not been factored in, the justification of attacks will be useful to compare with other visualizations for predictive analysis. 

```{r}
  fatal_race_table <- table(fatal_summary$Fatal, fatal_summary$Officer_Race)
barplot(fatal_race_table, legend.text = TRUE, beside = TRUE, cex.names = 0.15, xlab = "Fatality and Race", ylab = "Count", main = "Does Race Determine if an Interaction is Fatal?")
```

- As seen in the figure above, there are significantly more white officers reported in the data set than any other race. When looking at white officers, it can be seen that there is only slightly more encounters with non-fatal injuries than fatal injuries. However, when looking at officers of Hispanic or african american descent, there is significantly more fatal encounters than non-fatal. This can be used to show that for African Americans and Hispanic officers, they are more likely to be fatally injured which can used to prove racial injustice.

```{r}
fatal_gender_table <- table(fatal_summary$Fatal, fatal_summary$Officer_Gender)
barplot(fatal_gender_table, legend.text = TRUE, beside = TRUE, cex.names = 1, xlab = "Fatality and Gender", ylab = "Count", main = "Does Gender Determine if an Interaction is Fatal?")
```

- When looking at the fatality of police encounters with respect to gender, it can be seen that male officers are more often fatally injured whereas female officers are more often not. This can be used to show some slight gender inequality but seeing as there is very little disparity, many of these incidents can be concluded as not gender equality as well since the number of fatal and non-fatal encounters is very close for both genders.

